Strong lensing selection effects

### 2023-05-26 paper/seleff.tex
rewritten intro, magnification-dependent selection

   \author{Alessandro Sonnenfeld\inst{\ref{sjtu},\ref{leiden}}\and
           Shun-Sheng Li\inst{\ref{leiden}}\and
           Giulia Despali\inst{\ref{ita}}\and
           Anowar J. Shajib\inst{\ref{chicago},\ref{kicp},\ref{einstein}}\and
           Edward N. Taylor\inst{\ref{swin}}
          }

   \institute{
Department of Astronomy, School of Physics and Astronomy, Shanghai Jiao Tong University, Shanghai 200240, China\\
              \email{sonnenfeld@sjtu.edu.cn}\label{sjtu}
\and
Leiden Observatory, Leiden University, P.O. Box 9513, 2300 RA Leiden, The Netherlands\label{leiden}\and
Institut für Theoretische Astrophysik, Zentrum für Astronomie, Heidelberg Universität, Albert-Ueberle-Str. 2, 69120, Heidelberg, Germany\label{ita}\and
Department of Astronomy \& Astrophysics, University of Chicago,
Chicago, IL 60637, USA\label{chicago}\and
Kavli Institute for Cosmological Physics, University of Chicago,
Chicago, IL 60637, USA\label{kicp}\and
NHFP Einstein Fellow\label{einstein}\and
Centre for Astrophysics and Supercomputing, Swinburne University of Technology, Hawthorn 3122, Australia\label{swin}
             }

@@

Strong lenses, however, are a biased subset of the general population of galaxies and background sources.
A necessary condition for a galaxy to act as a strong lens with respect to a given source is that its projected surface mass density $\Sigma(\boldsymbol\theta)$ must be larger than the critical surface mass density for lensing $\Sigma_{\mathrm{cr}}$ at at least one position $\boldsymbol\theta$ \citep{SEF92}.
This condition excludes objects with a very diffuse mass distribution from the population of lenses.
In general, galaxies with a higher concentration of mass are more likely to be strong lenses, and are therefore over-represented in lens samples.
The distribution of strongly lensed sources can also be biased with respect to the general population of background galaxies: for instance, lensing magnification allows the detection of fainter objects with respect to the field.

In general, the probability distribution $\prsl$ of a sample of strong lenses from a given survey with selection criterion $S$ is given by \citep{Son22}:
\begin{equation}\label{eq:one}
\prsl(\psilens,\psisource|S) \propto \prlens(\psilens)\prsource(\psisource)\psel(\psilens,\psisource|S).
\end{equation}
In the equation above, $\psilens$ is the set of parameters describing the properties of foreground galaxies that are relevant for lensing, such as their redshift and mass distribution; $\psisource$ is the set of parameters describing background sources; $\prlens$ and $\prsource$ describe the parent distribution of foreground galaxies and background sources in the absence of lensing; and $\psel$ describes the probability of selecting a lens-source system with parameters $\psilens$ and $\psisource$ given the criterion $S$ used to define a lens.
This last factor takes into account both physical effects, that is whether a lens with parameters $\psilens$ produces a strongly lensed image of a source with parameters $\psisource$, and survey selection effects, that is whether such an image can be detected and recognised as a strong lens.

>>>>

   \author{Alessandro Sonnenfeld\inst{\ref{sjtu},\ref{leiden}}\and
           Shun-Sheng Li\inst{\ref{leiden}}\and
           Giulia Despali\inst{\ref{ita}}\and
           Raphael Gavazzi\inst{\ref{marseille}}\and
           Anowar J. Shajib\inst{\ref{chicago},\ref{kicp},\ref{einstein}}\and
           Edward N. Taylor\inst{\ref{swin}}
          }

   \institute{
Department of Astronomy, School of Physics and Astronomy, Shanghai Jiao Tong University, Shanghai 200240, China\\
              \email{sonnenfeld@sjtu.edu.cn}\label{sjtu}
\and
Leiden Observatory, Leiden University, P.O. Box 9513, 2300 RA Leiden, The Netherlands\label{leiden}\and
Institut für Theoretische Astrophysik, Zentrum für Astronomie, Heidelberg Universität, Albert-Ueberle-Str. 2, 69120, Heidelberg, Germany\label{ita}\and
CNRS and CNES, Laboratoire d'Astrophysique de Marseille, Aix-Marseille Universit\'{e} 38, Rue Fr\'{e}d\'{e}ric Joliot-Curie, F-13388 Marseille, France\label{marseille}\and
Department of Astronomy \& Astrophysics, University of Chicago,
Chicago, IL 60637, USA\label{chicago}\and
Kavli Institute for Cosmological Physics, University of Chicago,
Chicago, IL 60637, USA\label{kicp}\and
NHFP Einstein Fellow\label{einstein}\and
Centre for Astrophysics and Supercomputing, Swinburne University of Technology, Hawthorn 3122, Australia\label{swin}
             }

@@

Strong lenses, however, are a biased subset of the general population of galaxies and background sources.
In order for a lens-source pair to be included in a strong lensing survey, the following conditions must be met:
\begin{enumerate}
\item The lens must produce multiple images of the source
\item The multiple images must be detected and resolved
\item The system must be recognized as a lens by the survey.
\end{enumerate}
Condition 1 poses constraints on the mass distribution of the lens, as well as on the geometry of the system. 
First of all, not all lenses are capable of producing multiple images, but only those whose surface mass density $\Sigma(\boldsymbol\theta)$ is larger than the critical surface mass density for lensing $\Sigma_{\mathrm{cr}}$ at at least one position $\boldsymbol\theta$ \citep{SEF92}.
This condition excludes objects with a very diffuse mass distribution from the population of lenses.
Second, objects with a higher concentration of mass allow for a larger portion of the source plane to be multiply imaged, and therefore are over-represented in lens samples. Third, at fixed lens properties, sources that are located further away along the line of sight are more likely to be lensed.

%First, the projected surface mass density $\Sigma(\boldsymbol\theta)$ must be larger than the critical surface mass density for lensing $\Sigma_{\mathrm{cr}}$ at at least one position $\boldsymbol\theta$ \citep{SEF92}, a condition that excludes objects with a very diffuse mass distribution from the population of lenses.
%Second, the source must be sufficiently close in projection to the lens center, in a way that depends on the lens mass distribution. Galaxies with a higher concentration of mass allow for a larger portion of the source plane to be multiply imaged, and therefore are over-represented in lens samples. Third, at fixed lens properties, sources that

Condition 2 sets limits on the observable properties of the sources and, again, the lens mass distribution, in a way that depends on the survey characteristics.
On the one hand, brighter sources are more likely to be detected when lensed. On the other hand, lensing magnification allows the detection of fainter objects with respect to the field.
However, if the angular separation between the multiple images is too small, the system may not be classified as a lens.
Since the image separation increases with lens mass, this means that the lens sample is biased against low-mass objects.

Condition 3 can introduce additional biases, depending on the efficiency of the lens finding process on which the survey is based. Most commonly, lenses with a smaller image separation are more difficult to identify even if detected and resolved, because of contamination from the lens galaxy. This introduces an additional bias towards lenses with a larger or more concentrated mass.

%A necessary condition for a galaxy to act as a strong lens with respect to a given source is that its projected surface mass density $\Sigma(\boldsymbol\theta)$ must be larger than the critical surface mass density for lensing $\Sigma_{\mathrm{cr}}$ at at least one position $\boldsymbol\theta$ \citep{SEF92}.
%This condition excludes objects with a very diffuse mass distribution from the population of lenses.
%In general, galaxies with a higher concentration of mass are more likely to be strong lenses, and are therefore over-represented in lens samples.
%The distribution of strongly lensed sources can also be biased with respect to the general population of background galaxies: for instance, lensing magnification allows the detection of fainter objects with respect to the field.

The cumulative effect of these conditions can be summarised as follows.
%In general, 
The probability distribution $\prsl$ of a sample of strong lenses from a given survey with selection criterion $S$ is given by \citep{Son22}:
\begin{equation}\label{eq:one}
\prsl(\psilens,\psisource|S) \propto \prlens(\psilens)\prsource(\psisource)\psel(\psilens,\psisource|S).
\end{equation}
In the equation above, $\psilens$ is the set of parameters describing the properties of foreground galaxies that are relevant for lensing, such as their redshift and mass distribution; $\psisource$ is the set of parameters describing background sources; $\prlens$ and $\prsource$ describe the parent distribution of foreground galaxies and background sources in the absence of lensing; and $\psel$ describes the probability of selecting a lens-source system with parameters $\psilens$ and $\psisource$ given the criterion $S$ used to define a lens.
This last factor takes into account both physical effects, that is whether a lens with parameters $\psilens$ produces a strongly lensed image of a source with parameters $\psisource$, and survey selection effects, that is whether such an image can be detected and recognised as a strong lens.
